Automated Program Verification and Synthesis of Embedded \& Cyber-physical Systems: Advances and Challenges

### 2017-08-11 paper.tex
corrections

Timed automata targets real-time systems and employs finite automata with clocks, {\it i.e.}, real variables that follow continuous trajectories (behaviors) with constant slope. Hybrid automata bridges digital processing and analog entities and consequently generalizes timed automata, by also employing finite automata with real variables, whose trajectories follow more general dynamical laws~\cite{Henzinger95}.  Timed automata and hybrid automata are deterministic models for hybrid systems that built the basis of HS representations, which were subdivided into different classes of hybrid automata, depending on mapping functions ($r$ and $v$), {\it e.g.}, linear hybrid automata~\cite{Lafferriere99}, rectangular hybrid automata~\cite{Puri94,Henzinger95}, nonlinear hybrid automata~\cite{Henzinger95NL,Broucke1997}, integration graphs~\cite{Kesten1993}, and linear stochastic hybrid automata~\cite{Julius09}.

Finally, Rungger and Tabuada~\cite{Rungger16} and Tabuada {\it et al.}~\cite{Tabuada14} proposed a general and wide model for ECPSs, which is based on hybrid automata, by providing the concept of robustness and important tools for robust stability analysis.

@@

Given a class of ECPSs $\Omega$ and a set $\Phi$ of desired properties, a class of verification problems is called decidable if there is an algorithm that is able to decide if a finite number of steps $\forall \mathcal{O}\in\Omega$ satisfy every property $\mathcal{L}\in \Phi$~\cite{Alur00}. Decidability is a key issue in ECPS verification, since there are uncountable states and complex dynamics to be considered, {\it i.e.}, hybrid automata and their variations (usually employed to model ECPSs) are infinite-state machines~\cite{Henzinger95}. Generally, the properties that must hold for ECPSs are related to safety, liveness, and reachability. In particular, model-checking procedures have been employed to verify those properties even under undecidability conditions via approximations, {\it e.g.}, bounding the number of steps from the initial state, which leads to bounded model checking~\cite{Veanes09}.

In addition, Shoukry {\it et al.} \cite{Shoukry} tackled another important problem related to ECPSs: security. Indeed, the essence of ECPSs claims for sophisticated interfacing ({\it e.g.}, sensors and network connections), which normally leads to security issues~\cite{cpsattack}, given that wrong control actions can be generated through corrupted measurements. The mentioned authors proposed a methodology for estimating the state of physical systems, through formal methods, even if sensor inputs are contaminated with erroneous or intentionally corrupted data, in order to still be able to execute suitable control commands. Indeed, many situations can be regarded as satisfiability problems and solved, for instance, with approaches based on Satisfiability Modulo Theories (SMT)~\cite{Araujo16} \textcolor{red}{eh bom citar mais fontes}, which indicates many links with other research areas and might lead to a unified approach regarding system verification, security enforcement, and interaction handling.

@@

>>>>

Timed automata targets real-time systems and employs finite automata with clocks, {\it i.e.}, real variables that follow continuous trajectories (behaviors) with constant slope. Hybrid automata bridges digital processing and analog entities and consequently generalizes timed automata, by also employing finite automata with real variables, but whose trajectories follow more general dynamical laws~\cite{Henzinger95}.  Timed automata and hybrid automata are deterministic models for hybrid systems that built the basis of HS representations, which were subdivided into different classes of hybrid automata, depending on mapping functions ($r$ and $v$), {\it e.g.}, linear hybrid automata~\cite{Lafferriere99}, rectangular hybrid automata~\cite{Puri94,Henzinger95}, nonlinear hybrid automata~\cite{Henzinger95NL,Broucke1997}, integration graphs~\cite{Kesten1993}, and linear stochastic hybrid automata~\cite{Julius09}. Recently, Rungger and Tabuada~\cite{Rungger16} and Tabuada {\it et al.}~\cite{Tabuada14} proposed a general and wide model for ECPSs, which is based on hybrid automata, by providing the concept of robustness and important tools for robust stability analysis. In addition to HA and TA based models, there are a number of other ECPS modeling initiatives based on (exact or approximate) bisimulation~\cite{Girard11}, simulation~\cite{Rungger13} and interval analysis~\cite{Li16}.

Another approach to ECPS modeling tackles quantized control systems and finite-word length (FWL) effects on the performance of those systems. Differently from bisimulation, quantized models are built from a countable subset of the (quantized) input space~\cite{Tabuada07}. In particular, Keel and Bhattacharyya showed that even robust and optimal controllers may still present an undesired property~\cite{Keel01,bhattacharyya}: fragility or sensitivity to implementation aspects ({\it e.g.}, FWL representations of digital controllers). Therefore, various studies tried link sampled-data theory to hybrid-system models~\cite{Bicchi02,Petreczky10,Petreczky07,Ye98}.

Finally, in addition to ECPS-dynamics modeling, it is often interesting to establish invariants on a ECPS's state space, which can be used for accelerating formal verification and synthesis processes regarding those systems. Such invariants might be synthesized, in order to point out state-space boundaries~\cite{BenSassi12} and determine ECPS stability~\cite{Paul14} or reachability~\cite{Li16_2,Li16}.

@@

Given a class of ECPSs $\Omega$ and a set $\Phi$ of desired properties, a class of verification problems is called decidable if there is an algorithm that is able to decide if a finite number of steps $\forall \mathcal{O}\in\Omega$ satisfy every property $\mathcal{L}\in \Phi$~\cite{Alur00}. Decidability is a key issue in ECPS verification, since there are uncountable states and complex dynamics to be considered, {\it i.e.}, hybrid automata and their variations (usually employed to model ECPSs) are infinite-state machines~\cite{Henzinger95}. In particular, the safety verification of ECPS is undecidable, except in very specific cases as in systems that can be represented by timed automata and rectangular automata~\cite{Alur11}. An alternative to ensure that ECPS trajectories will not pass through unsafe states is the use of barrier certificates~\cite{Prajna07,Prajna04}. An example of barrier certificate is the stability, that ensures the convergence of states to an equilibrium point and consequently that there exists a region from where any trajectory remains inside of this region. The stability may be ensured based on Lyapunov theory (for a wide class of model abstractions)~\cite{tabuada2009verification}, or linear system stability theory for discrete simulations~\cite{Bessa17}.

Generally, properties that must hold for ECPSs are related to safety/reachability and liveness. Therefore, model-checking procedures have been employed to verify those properties, even under undecidability conditions via approximations, {\it e.g.}, bounding the number of steps from initial states, which leads to bounded model checking (BMC)~\cite{Veanes09}.

Along with CPS modeling and verification techniques purely based on HA, one can also highlight logic-based approaches~\cite{Phan15} that may still use HA, but are not based on a algebraic description of states and transitions. Some examples of logic-based verification are theorem proving techniques, {\it e.g.} deductive verification (i.e., interactive theorem proving)~\cite{Nipkow02} and automatic theorem proving~\cite{Platzer07}, and reduction-based techniques, among which BMC stands out~\cite{handbook09}.

%Much of the work related to the formal verification of ECPS seeks to ensure the safety, robustness, and reachability properties of an HA. 
Recently, some studies have addressed ECPS formal verification considering discretization, switching, and quantization effects.  Dugirala and Viswanathan~\cite{Duggirala2015} verified embedded control software considering real time issues, {\it e.g.}, delayed responses and possible loss of real time deadlines. Anta {\it et al.}~\cite{Anta10} presented a tool called Costan, which finds errors in mathematical-model implementations and verifies whether error is tolerated, considering quantization effects and fixed-point implementation, while focusing on quantization error effects over system stability. Similarly, Ismail {\it et al.}~\cite{dsv_spin2015} preseted DSVerifier, a BMC tool for digital systems that employs SMT/SAT solvers as back-ends, in order to provide support for digital system design and verification considering FWL effects. In particular, DSVerifier checks if a designed digital controller and/or closed-loop ECPS presents the desired performance, when it is implemented with a FWL format~\cite{Bessa16,Bessa17}.

Finally, it is worth mentioning that there are other alternatives for ECPS modeling and analysis, which are not based on verification, but instead on simulators~\cite{Phan2014,Nakajima12,Junjie12}. Although such approaches generally fail in ensuring that some properties hold, they present good scalability, which is ideal for large scale and complex systems that are hardly modeled through automata- or logig-based approaches. Modelica~\cite{Junjie12} and Simulink/MATLAB~\cite{Nakajima12} are important examples of widely used simulators for ECPSs. In order to achieve a balanced trade-off between ECPS-simulation scalability and correctness of verification methods, statistical model checking have been employed to formally verify ECPSs~\cite{Clarke11}.

In addition, Shoukry {\it et al.} \cite{Shoukry} tackled another important problem related to ECPSs: security. Indeed, the essence of ECPSs claims for sophisticated interfacing ({\it e.g.}, sensors and network connections), which normally leads to security issues~\cite{cpsattack}, given that wrong control actions can be generated through corrupted measurements. The mentioned authors proposed a methodology for estimating physical-system state, through formal methods, even if sensor inputs are contaminated with erroneous or intentionally corrupted data, in order to still be able to execute suitable control commands. Cyber-security is also a concern of Fiore {\it et al.}, who aim to ensure secure-state estimation of UAV systems under sparse attacks~\cite{Fiore17}. Indeed, many situations can be regarded as satisfiability problems and solved, for instance, with approaches based on Satisfiability Modulo Theories (SMT)~\cite{Araujo16,TrindadeDAES2016,Rahman13}, which indicates many links with other research areas and might lead to a unified approach regarding system verification, security enforcement, and interaction handling.

@@

Control synthesis for ECPSs consists in finding a controller or an invariant control law $\mathcal{C}$, such that a closed-loop system composed of $(\mathcal{C},\Omega)$ satisfies every property $\mathcal{L}\in \Phi$.

An efficient way to synthesize reliable ECPSs is the Counterexample Guided Inductive Synthesis (CEGIS)~\cite{David15}. Its key idea is the definition (by a user) of a set of safe states, {\it i.e.}, the specification of $\Phi$, as well as a generic template for a control system that must satisfy such a specification. The, a CEGIS engine compute values for the unknown symbols of the chosen template, which hold $\Phi$~\cite{RienerKFB16}. Abate~{\it et al.}~\cite{Abate17,abatecav2017} presented a method for synthesizing stable controllers that are suitable to continuous plants given as transfer functions~\cite{Abate17} and state-space representations~\cite{abatecav2017}, which exploits bit-accurate verification of software implemented in digital microcontrollers~\cite{Bessa16,Bessa17,dsv_spin2015}.

% The mentioned authors developed a tool called DSSynth, which is able to synthesize digital controllers that are algorithmically and numerically sound, w.r.t. stability. DSSynth marks the first use of counterexample-guided inductive synthesis~\cite{sketch} to synthesize digital controllers, considering physical plants with uncertain models and FWL effects; however, low-level implementation errors ({\it e.g.}, limit cycles) are not further investigated.
